The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-12 paper/iphas-dr2.tex
Update iphas-dr2.tex
I've been through everything now, except the figure captions (which I probably will leave to you).  Note that I've numbered the Sh 2-82 candidate Ha emitters but don't know which two are the objects in common with Yu+Wang (over to you to resolve the 'x and y' now in the text).  In section 5, you'll find a couple of comments, that begin with GEERT that you need to address.  Also you now have a Parker05 reference to add to the bibliography.  ...still to do: expand section 8.

Having obtained a quality-checked 
and re-calibrated data set, 
we now turn to the problem
of transforming the observations 
into a user-friendly catalogue.
The aim of this catalogue is to detail
the best-available information for each unique source
in a convenient format,
including flags to warn about quality issues 
such as source blending and saturation.
Compiling the catalogue essentially required four steps:
\begin{enumerate}
\item the single-band detection tables 
produced by the CASU pipeline 
were augmented with new columns
and warning flags;
\item the detection tables were merged into multi-band field catalogues;
\item the overlap regions of the field catalogues 
were cross-matched to flag duplicate measurements 
and identify the primary (best) detection 
of each unique source; and
\item these primary detections
were compiled into the final source catalogue.
\end{enumerate}
Each of these four steps are explained next.

@@

As the first step, the detection tables 
were enhanced by creating new columns.
This is necessary because the tables 
generated by the CASU pipeline 
summarise the detections 
in their original CCD units,
e.g. source positions are given in pixel coordinates 
and photometry in number counts.
To transform these measurements into
user-friendly fields,
we have largely adopted the units and naming conventions
which are in use at the 
WFCAM Science Archive \citep[WSA;][]{Hambly2008}
and the VISTA Science Archive \citep[VSA;][]{Cross2012}.
These archives curate the near-infrared data from both
the UKIDSS Galactic Plane Survey \citep[GPS;][]{Lucas2008}
and the 
VISTA Variables in the Via Lactea survey \cite[VVV;][]{Minniti2010}.
Both these surveys provide high-resolution JHK photometry
in the Galactic Plane.
There is a significant degree of overlap
between the footprints of UKIDSS/GPS and IPHAS,
and hence by adopting a similar catalogue format
we hope to encourage scientific applications
which combine both data sets.

@@

Firstly, we note that each source is uniquely identified by an
IAU-style designation of the form ``IPHAS2\ JHHMMSS.ss+DDMMSS.s''
(cf. column \emph{name} in Appendix~\ref{app:columns}),
where ``IPHAS2'' refers to the present
data release and the remainder of the string
denotes the J2000 coordinates in sexagesimal format.
For convenience, the coordinates
are also included in decimal degrees
(columns \emph{ra} and \emph{dec})
and in Galactic coordinates
(columns \emph{l} and \emph{b}).
We have also included an internal object identifier string 
of the form ``\#run-\#ccd-\#detection''
(e.g. ``64738-3-6473''),
which documents the INT exposure number (\#run),
the CCD number (\#ccd),
and the row number in the CASU detection table (\#detection)
-- column names \emph{rDetectionID},
\emph{iDetectionID}, \emph{haDetectionID}.

Photometry is provided based on the 2.3-arcsec diameter circular aperture
by default (columns \emph{r}, \emph{i}, \emph{ha}).
The choice of this aperture size as the default 
is based on a trade-off between concerns 
about small number statistics and centroiding errors
for small apertures on one hand,
and diminishing signal-to-noise ratios and source confusion
for large apertures on the other hand.
The user is not restricted to this choice, because
the catalogue also provides magnitudes
using three alternative aperture sizes:
the peak pixel height 
(columns \emph{rPeakMag}, \emph{iPeakMag}, \emph{haPeakMag}),
the circular 1.2-arcsec-diameter aperture 
(\emph{rAperMag1}, \emph{iAperMag1},
 \emph{haAperMag1}) and
the 3.3-arcsec-diameter aperture 
(\emph{rAperMag3}, \emph{iAperMag3},
 \emph{haAperMag3}).

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude.
Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error,
and we do not recommend its use.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band, and hence the $(r-i)$ and $(r-\ha)$ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
Such very bright stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
We will explain below that no such detections 
have been included in the catalogue.
An alternative detection was available in essentially all these situations
as a result of the IPHAS field pair strategy,
and hence it was not necessary to include these internal warning flags
in the source catalogue.

@@

The second step in compiling the source catalogue
is to merge the contemporaneous trios
of $r$, $i$, \ha\ detection tables
into multi-band field catalogues.
This required a positional matching procedure 
to link sources between the three bands
based on their position on the sky.
We used the \textsc{tmatchn} function 
of the \textsc{stilts} software for this purpose,
which allows rows from multiple tables to be matched \citep{Taylor2006}.
The result of the procedure is a band-merged catalogue
in which each row corresponds to a group of linked $r$, $i$, and \ha\ detections
which satisfy a maximum matching distance criterion in a pair-wise sense.
Sources for which no counterpart was identified
are retained in the catalogue as single-band detections.

We employed a maximum matching distance of 1~arcsec,
which was chosen based on a trade-off between 
completeness and reliability.
On one hand, a matching distance larger than 1~arcsec 
was found to allow too many spurious and unrelated sources 
to be linked. 
On the other hand, a value smaller than 1~arcsec 
would pose problems for very faint sources 
with large centroiding errors, 
and would occasionally fail near CCD corners,
where the astrometric solution can 
show systematic errors which exceed 0.5~arcsec.
The position offsets between the $r$ and $i$, \ha\ detections
have been included in the catalogue 
and can hence be tightened by the user if necessary
(columns \emph{iXi}, \emph{iEta}, \emph{haXi}, \emph{haEta}).
We note that UKIDSS/GPS adopted 
the same maximum matching distance of 1 arcsec
for similar reasons \citep{Hambly2008},
and warn that applications which require absolute
astrometry with sub-arcsecond accuracy must
use these catalogues with caution.

@@

We explained earlier that the survey contains repeat observations
of identical sources as a result of overlaps in the data-taking pattern.
Amongst all sources in the magnitude range $13<r<19$,
we find that 65\% were detected twice and 25\% were detected three times or more.
Only 9\% were detected once.
Unsurprisingly, their spatial distribution traces
the inter-CCD gaps and footprint edges.

The principal aim of this data release is to provide 
reliable photometry at a single epoch,
and hence we have decided
to focus on providing the magnitudes
and coordinates using only the \emph{best-available} 
detection of each object -- 
hereafter referred to as the \emph{primary} detection.
Although overlapping fields could have been co-added 
to gain a small improvement in depth, 
we have decided against this for two reasons.
Firstly, combining the information from multiple epochs
would make the photometry of variable stars difficult to interpret.
Secondly, co-adding would cause the image quality to degrade towards the mean,
which is a draw-back for crowded fields.

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was observed 
within 10 minutes of the primary,
we have included the identifier and the default photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Such a secondary detection is available for 66\% of the sources brighter than $r<20$
due to the IPHAS field pair strategy.
In addition, the complete set of detection tables -- one for each exposure -- 
is made accessible on our website to allow other uses of the data.
A user-friendly catalogue of secondary detections 
has not been compiled at present 
but may be part of a future data release.

The primary detection is defined as the entry in the 
set of band-merged field catalogues which provides 
the most reliable information for a unique source.
Primary detections have been selected using a so-called \emph{seaming} procedure,
which we adapted from the algorithm developed for the WSA\footnote{http://surveys.roe.ac.uk/wsa/dboverview.html\#merge}.
In brief, the first step of the procedure is to identify all the duplicate detections
by cross-matching the overlap regions of all field catalogues,
again using a maximum matching distance of 1\arcsec.
The duplicate detections for each unique source are then ranked according to
(i) filter coverage, (ii) quality score (column \emph{errBits}),
and (iii) the average seeing of stars in the CCD frame rounded to 0.2~arcsec.
If this ranking scheme reveals multiple `winners' of identical quality,
then the one that was observed closest to the optical axis of the camera is chosen.

@@

As the final step, the primary detections
selected above were compiled
into the 98-column source catalogue
that is described in Appendix~\ref{app:columns}
and is made available on our website and through Vizier.
The entire list of sources naturally includes 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20\%;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs,
i.e. we require the \emph{errBits} quality score
to be smaller than 64.
\end{enumerate}

A total of 219 million primary detections satisfied
the above criteria and have been included in the catalogue.
Amongst these, 158 million objects 
are detected in all three bands (72\%),
30 million are detected in two bands (14\%),
and 31 million entries are single-band detections (14\%).
Half of the single-band detections were made in the $i$-band.
This is likely explained by the fact that $i$ is least
affected by interstellar extinction and can occasionally pick up
highly-reddened objects which are otherwise lost in $r$/\ha.

@@

Having explained how the catalogue was created,
we now offer an overview of its properties
by discussing  
(i) the recommended quality criteria,
(ii) the photometric uncertainties and repeatability,
and (iii) the source densities and the frequency of source blending.

@@

Like any other photometric survey,
the majority of the objects in our catalogue
are faint sources observed near the detection limits:
55\% of the entries in the catalogue
are fainter than $r > 20$.
%and 18\% are even fainter than $r > 21$.
The measurements of faint objects
are naturally prone to larger
random and systematic uncertainties,
for example, an inaccurately subtracted background
will introduce a proportionally larger systematic error
for a faint object.
Most scientific applications will hence require a set of
quality criteria to be enforced for the purpose
of removing faint and low-quality objects.

The choice of quality criteria is a complicated
trade-off between completeness on one hand
and accuracy on the other.
To aid users we have listed two sets of
recommended quality criteria 
in Tables~\ref{tab:reliable} and \ref{tab:veryreliable}.

Firstly, Table~\ref{tab:reliable} details
a set of minimum quality criteria
which should benefit most applications
which require reliable $(r-i)$ and $(r-\ha)$ colours
without removing more than 80\%
of the sources brighter than $r < 19$.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2.3-arcsec diameter
aperture magnitude is inconsistent 
with its alternative 1.2-arcsec diameter measurement.
This last criterion is a proxy
for identifying objects which are affected
by inaccurate background subtraction
or failed source deblending.
A total of 86 out of 219 million sources 
(39\%) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as ``reliable''.
For convenience, the catalogue contains a boolean column
named \emph{reliable} which flags these objects
and makes their selection easy.

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distribution for objects with and without the 
\emph{reliable} and \emph{veryReliable} criteria applied. 
We find that 81\% of the sources 
in the magnitude range $13 < r < 19$
are considered \emph{reliable},
which drops to 72\% in the range $19 < r < 20$
and 9\% at $r>20$.
In contrast, only 54\% of the sources 
in the magnitude range $13 < r < 19$
are considered \emph{veryReliable}.
In \S\ref{sec:densities} we will show that these
stricter criteria remove a lot
of objects at early Galactic longitudes 
where source blending can sometimes affect
more than a quarter of the objects.
The \emph{veryReliable} flag should hence
only be used in applications which require very reliable photometry
at the expense of completeness.

@@

It is easy to see how the quality criteria
may be adapted to be more tolerant.
For example, by raising the allowed photometric uncertainties
from 0.1 mag to 0.2 mag,
42 million additional candidate sources can be added.

@@

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude in each band.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$/\ha\ for the default 2.3\arcsec\ aperture.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for $(r-i)$ and $0.44\pm0.03$ for $(r-\ha)$.
The better depth of $r$ is hence compensated
by the fact that stars tend to have 
brighter magnitudes in both $i$ and \ha.

@@

The number of sources in our catalogue,
counted in 1-by-10~deg$^2$ strips,
is shown in Fig.~\ref{fig:sourcecount}
as a function of Galactic longitude (thick blue line).	
Unsurprisingly, we find the number of sources
to increase towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 300\,000 objects per square degree,
which is five times more than the density
found near $l\simeq 180^\circ$.

In addition to the global trend,
variations are also apparent on smaller scales.
For example, we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction
\citep[known as `the Great Rift', e.g.][]{BokBok}.

We warn that the source counts reported here 
have not been corrected for survey completeness
or variations in the depth of different survey fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).
In a forthcoming paper,
we will calibrate the IPHAS-based source densities
by injecting artificial stars into the images
and measuring their recovery rates
(Farnhill et al., in preparation).
Such calibrated density maps may be used 
to constrain detailed models of our Galaxy.

@@

In Fig.~\ref{fig:sourcecount} we also shows the number
of non-blended sources (thin red line).
These are sources for which the \emph{deblend} flag is {\sc false},
i.e. sources for which the CASU pipeline did not have to apply 
a deblending procedure to separate the flux
originating from two or more overlapping objects.
We find that the fraction of sources affected by blending problems
is strongly correlated with the local source density.
For example, only 11\% of the sources are blended
at $l>90^\circ$, whereas 24\% are blended at $l<90^\circ$.

As we explained earlier, blended sources must be used with caution.
In particular, we warn that blended objects
are likely to have fallen victim
to source confusion during the band-merging procedure.
We have started investigating the use of
more advanced PSF-fitting routines
in which sources are measured simultaneously across all bands,
which may be guided by an external astrometric catalogue
such as the one that is to be provided by Gaia.

@@

We conclude this paper by demonstrating how the unique
$(r-\ha,\ r-i)$ colour-colour diagram offered by this catalogue
can readily be used to
(i) characterise the extinction regime at different sight-lines, and
(ii) identify \ha\ emission-line objects.

@@

The survey's unique $(r-\ha)$ colour,
when combined with $(r-i)$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
Put differently, the main sequence in the $(r-\ha,\ r-i)$ diagram
runs in a direction that is different from the reddening vector,
because the ($r-\ha$) colour tends to act
as a coarse proxy for spectral type
and is less sensitive to reddening than $(r-i)$.
As a result, the distribution of a stellar population
in the IPHAS colour-colour diagram
can offer a handle on the properties of the population
and the extinction along a line of sight.

This is demonstrated 
in Fig.~\ref{fig:l180}, \ref{fig:l45} \& \ref{fig:l30},
where we present three sets of IPHAS colour/magnitude diagrams
towards three distinct sight-lines
located at Galactic longitudes 
$180^\circ$, $45^\circ$ and $30^\circ$, respectively.
Each figure contains all the sources
which are flagged as \emph{veryReliable}
and are located in a region of one square degree 
centred on the coordinates indicated in the diagram
(i.e. within a radius of $0.564^\circ$ from the indicated sight-line).
For clarity, we have imposed the additional criterion
that the photometric uncertainties
must be smaller than 0.05 mag in each band,
corresponding to a magnitude limit near $\sim$19th magnitude.

>>>>

Having obtained a quality-checked 
and re-calibrated data set, 
we now turn to the problem
of transforming the observations 
into a user-friendly catalogue.
The aim is to present
the best-available information for each unique source
in a convenient format,
including flags to warn about quality issues 
such as source blending and saturation.
Compiling the catalogue involved four steps:
\begin{enumerate}
\item the single-band detection tables 
produced by the CASU pipeline 
were augmented with new columns
and warning flags;
\item the detection tables were merged into multi-band field catalogues;
\item the overlap regions of the field catalogues 
were cross-matched to flag duplicate measurements 
and identify the primary (best) detection 
of each unique source; and
\item these primary detections
were compiled into the final source catalogue.
\end{enumerate}
Each of these four steps are explained next.

@@

Enhancement of the detection tables by creating new columns
is the necessary first step because the tables generated by the CASU pipeline
refer e.g. to source positions in pixel coordinates, to photometric measures 
in number counts etc., rather than in directly useful astronomical terms.
To transform these data into user-friendly quantities,
we have largely adopted the units and naming conventions
which are in use at the 
WFCAM Science Archive \citep[WSA;][]{Hambly2008}
and the VISTA Science Archive \citep[VSA;][]{Cross2012}.
These archives curate the high-resolution near-infrared photometry from both
the UKIDSS Galactic Plane Survey \citep[GPS;][]{Lucas2008}
and the 
VISTA Variables in the Via Lactea survey \cite[VVV;][]{Minniti2010}.
There is a significant degree of overlap
between the footprints of UKIDSS/GPS and IPHAS,
and hence by adopting a similar catalogue format
we hope to facilitate scientific applications
which combine both data sets.

@@

First, we note that each source is uniquely identified by an
IAU-style designation of the form ``IPHAS2\ JHHMMSS.ss+DDMMSS.s''
(cf. column \emph{name} in Appendix~\ref{app:columns}),
where ``IPHAS2'' refers to the present
data release and the remainder of the string
denotes the J2000 coordinates in sexagesimal format.
For convenience, the coordinates
are also included in decimal degrees
(columns \emph{ra} and \emph{dec})
and in Galactic coordinates
(columns \emph{l} and \emph{b}).
We have also included an internal object identifier string 
of the form ``\#run-\#ccd-\#detection''
(e.g. ``64738-3-6473''),
which documents the INT exposure number (\#run),
the CCD number (\#ccd),
and the row number in the CASU detection table (\#detection).
These columns are named \emph{rDetectionID},
\emph{iDetectionID}, \emph{haDetectionID}.

Photometry is provided based on the 2.3-arcsec diameter circular aperture
by default (columns \emph{r}, \emph{i}, \emph{ha}).
The choice of this aperture size as the default 
is based on a trade-off between concerns 
about small number statistics and centroiding errors
for small apertures on one hand,
and diminishing signal-to-noise ratios and source confusion
for large apertures on the other hand.
The user is not restricted to this choice, because
the catalogue also provides magnitudes
using three alternatives:
the peak pixel height 
(columns \emph{rPeakMag}, \emph{iPeakMag}, \emph{haPeakMag}),
the circular 1.2-arcsec-diameter aperture 
(\emph{rAperMag1}, \emph{iAperMag1},
 \emph{haAperMag1}) and
the 3.3-arcsec-diameter aperture 
(\emph{rAperMag3}, \emph{iAperMag3},
 \emph{haAperMag3}).

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude in
$r$.  Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error,
and shoud be viewed as indicative rather than as precise measurement.
%and we do not recommend reliance on it.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band, and hence the $(r-i)$ and $(r-\ha)$ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within a radius of 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
These brightest stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
No such detections have had to be included in the catalogue, as 
alternative detections were available in essentially all these situations
thanks to the IPHAS field pair strategy. In turn, this has meant
that it has not proved necessary to include these internal warning flags
in the pubished source catalogue.

@@

The second step in compiling the source catalogue
is to merge the contemporaneous trios
of $r$, $i$, \ha\ detection tables
into multi-band field catalogues.
This required a position matching procedure 
to link sources between the three bands.
We used the \textsc{tmatchn} function 
of the \textsc{stilts} software for this purpose,
which allows rows from multiple tables to be matched \citep{Taylor2006}.
The result of the procedure is a band-merged catalogue
in which each row corresponds to a group of linked $r$, $i$, and \ha\ detections
which satisfy a maximum matching distance criterion in a pair-wise sense.
Sources for which no counterpart was identified
are retained in the catalogue as single-band detections.

We employed a maximum matching distance of 1~arcsec,
which was chosen, trading off completeness against reliability.
On the one hand, a matching distance larger than 1~arcsec 
was found to allow too many spurious and unrelated sources 
to be linked. 
On the other, a value smaller than 1~arcsec 
would pose problems for very faint sources 
with large centroiding errors, 
and would occasionally fail near CCD corners,
where the astrometric solution can 
show systematic errors which exceed 0.5~arcsec.
The position offsets between the $r$ detection and detections in $i$, and/or \ha\
have been included in the catalogue, giving the user the option to tighten them 
further if necessary
(columns \emph{iXi}, \emph{iEta}, \emph{haXi}, \emph{haEta}).
We note that UKIDSS/GPS adopted 
the same maximum matching distance of 1 arcsec
for similar reasons \citep{Hambly2008},
and warn that applications which require absolute
astrometry with sub-arcsecond accuracy must
use these catalogues with caution.

@@

We explained earlier that the survey contains repeat observations
of identical sources as a result of field offsetting and overlaps.
Amongst all sources in the magnitude range $13<r<19$,
we find that 65\% were detected twice and 25\% were detected three times or more.
Only 9\% were detected once.
%Unsurprisingly, their spatial distribution traces
%the inter-CCD gaps and footprint edges.

Since the principal aim of this data release is to provide 
reliable photometry at a single epoch,
we have focused on providing the magnitudes
and coordinates from the \emph{best-available} 
detection of each object -- 
hereafter referred to as the \emph{primary} detection.
Although overlapping fields could have been co-added 
to gain a small improvement in depth, 
we have decided against this for two reasons.
Firstly, combining the information from multiple epochs
would make the photometry of variable stars difficult to interpret.
Secondly, co-adding would cause the image quality to degrade towards the mean,
which is particularly a draw-back for crowded fields.

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was collected 
within 10 minutes of the primary,
we have included the identifier and the default photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Such a secondary detection is available for 66\% of the sources brighter than $r<20$
due to the IPHAS field pair strategy.
In addition, the complete set of detection tables -- one for each exposure -- 
is made accessible on our website to allow other uses of the data.
%A user-friendly catalogue of secondary detections 
%has not been compiled at present 
%but may be part of a future data release.

%The primary detection is defined as the entry in the 
%set of band-merged field catalogues which provides 
%the most reliable information for a unique source.
Primary detections have been selected from all available detections using a so-called \emph{seaming} procedure,
which we adapted from the algorithm developed for the WSA\footnote{http://surveys.roe.ac.uk/wsa/dboverview.html\#merge}.
In brief, the first step is to identify all the duplicate detections
by cross-matching the overlap regions of all field catalogues,
again using a maximum matching distance of 1\arcsec.
The duplicate detections for each unique source are then ranked according to
(i) filter coverage, (ii) quality score (column \emph{errBits}),
and (iii) the average seeing of stars in the CCD frame rounded to 0.2~arcsec.
If this ranking scheme reveals multiple `winners' of seemingly identical quality,
then the one that was observed closest to the optical axis of the camera is chosen.

@@

As the final step, the primary detections
selected above were compiled
into the 98-column source catalogue
that is described in Appendix~\ref{app:columns}
and is made available on our website and through Vizier.
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20\%;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs,
i.e. we require the \emph{errBits} quality score
to be smaller than 64.
\end{enumerate}

A total of 219 million primary detections satisfied
the above criteria and have been included in the catalogue.
Amongst these, 158 million objects 
are detected in all three bands (72\%),
30 million are detected in two bands (14\%),
and 31 million entries are single-band detections (14\%).
Half of the single-band detections were made in the $i$-band.
This is because $i$ is least
affected by interstellar extinction and can occasionally pick up
highly-reddened objects which are otherwise lost in $r$ and \ha.

@@

%Having explained how the catalogue was created,
We now offer an overview of the properties of the
catalogue by discussing  
(i) the recommended quality criteria,
(ii) the photometric uncertainties and repeatability,
and (iii) the source densities and the frequency of source blending.

@@

Like any other photometric survey,
the majority of the objects in our catalogue
are faint sources observed near the detection limits:
55\% of the entries in the catalogue
are fainter than $r > 20$.
%and 18\% are even fainter than $r > 21$.
The measurements of faint objects
are naturally prone to larger
random and systematic uncertainties:
for example, an inaccurately-subtracted background
will introduce a proportionally larger systematic error
for a faint object.
Most scientific applications will hence require a set of
quality criteria to be enforced for the purpose
of removing lower-quality objects.

The choice of quality criteria will always tension 
completeness against accuracy.
To aid users we have listed two sets of
recommended quality criteria 
in Tables~\ref{tab:reliable} and \ref{tab:veryreliable}.

Table~\ref{tab:reliable} specifies
a set of minimum quality criteria
which should benefit most applications
which require reliable $(r-i)$ and $(r-\ha)$ colours
(GEERT: the rest of the sentence looks very odd to me -- I can't rephrase
because I'm not sure what you are trying to say) without removing more 
than 80\%
of the sources brighter than $r < 19$.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2.3-arcsec diameter
aperture magnitude is inconsistent 
with its alternative 1.2-arcsec diameter measurement.
This last criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed source deblending.
A total of 86 out of 219 million sources 
(39\%) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as ``reliable''.
For convenience, the catalogue contains a boolean column
named \emph{reliable} that directly flags these objects.

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distribution for objects with and without the 
\emph{reliable} and \emph{veryReliable} criteria applied. 
We find that 81\% of the sources 
in the magnitude range $13 < r < 19$
are considered \emph{reliable},
which drops to 72\% in the range $19 < r < 20$
and 9\% at $r>20$.
In contrast, only 54\% of the sources 
in the magnitude range $13 < r < 19$
are considered \emph{veryReliable}.
In \S\ref{sec:densities} we will show that these
stricter criteria remove large numbers
of objects at early Galactic longitudes 
where source blending can sometimes affect
more than a quarter of the objects.
The \emph{veryReliable} flag should hence
only be used in applications in which 
completeness is not a concern.

@@

It is easy to see how the quality criteria
may be adapted to be more tolerant.
For example, by raising the allowed photometric uncertainties
from 0.1 mag to 0.2 mag,
42 million candidate sources are added to the 109 
million satisfying the tighter error bound (GEERT - is this what you mean?).

@@

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude in each band.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$/\ha\ for the default 2.3\arcsec\ aperture.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for $(r-i)$ and $0.44\pm0.03$ for $(r-\ha)$. The
fainter limiting magnitude in $r$ is hence compensated
by the fact that most sources have 
brighter magnitudes in both $i$ and \ha.

@@

The number of sources in our catalogue,
counted in 1-by-10~deg$^2$ constant-longitude strips,
is shown in Fig.~\ref{fig:sourcecount}
as a function of Galactic longitude (thick blue line).	
Unsurprisingly, we find the number of sources
to increase towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 300\,000 objects per square degree,
which is six times more than the density
found near $l\simeq 180^\circ$.

In addition to the global trend,
real variations are also apparent on smaller scales.
For example, we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction
\citep[the extremities of `the Great Rift', e.g.][]{BokBok}.
However, we warn that the source counts shown 
have not been corrected for field pairs that have yet
to be released or for variations in the depth across the included fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).

In a forthcoming paper, properly calibrated maps of stellar density
of the northern Galactic Plane will be presented (Farnhill
et al., in preparation).  This will incorporate completeness
corrections based on the statistics of artificial source
recovery.  Such maps are of interest as tests of detailed models of 
our Galaxy.

@@

Fig.~\ref{fig:sourcecount} also shows the number counts
for non-blended sources (thin red line).
These are sources for which the \emph{deblend} flag is {\sc false},
i.e. sources for which the CASU pipeline did not have to apply 
a deblending procedure to separate the flux
originating from two or more overlapping objects.
This provides some insight into how the fraction of 
blend-affected sources correlates with the local source 
density.  In hard numbers, only 11\% of the sources are blended
at $l>90^\circ$, whereas 24\% are blended at $l<90^\circ$.

%As we explained earlier, blended sources must be used with caution.
Finally, we warn that blended objects
are more likely than unblended objects to have fallen victim
to source confusion during the band-merging procedure.  It is
important to bear this in mind when appraising stars of seemingly
unusual colours (such as candidate emission line stars) -- if 
blending is flagged, the probability that the unusual colour is 
spurious is greatly increased.
%We have started investigating the use of
%more advanced PSF-fitting routines
%in which sources are measured simultaneously across all bands,
%which may be guided by an external astrometric catalogue
%such as the one that is to be provided by Gaia.

@@

We conclude this paper by demonstrating how the unique
$(r-\ha,\ r-i)$ colour-colour diagram offered by this catalogue
can readily be used to
(i) characterise the extinction regime along different sight-lines, and
(ii) identify \ha\ emission-line objects.

@@

The survey's unique $(r-\ha)$ colour,
when combined with $(r-i)$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
Put differently, the main sequence in the $(r-\ha,\ r-i)$ diagram
runs in a direction that is at a large angle relative to the reddening vector,
because the ($r-\ha$) colour tends to act
as a coarse proxy for spectral type
and is less sensitive to reddening than $(r-i)$.
As a result, the distribution of a stellar population
in the IPHAS colour-colour diagram
can offer a handle on the properties of the population
and the extinction along a line of sight.

This is demonstrated 
in Fig.~\ref{fig:l180}, \ref{fig:l45} \& \ref{fig:l30},
where we present three sets of IPHAS colour/magnitude diagrams
towards three distinct sight-lines
located at Galactic longitudes 
$180^\circ$, $45^\circ$ and $30^\circ$, respectively.
Each figure contains all the sources
flagged as \emph{veryReliable} within a region of one square degree 
centred on the coordinates indicated in the diagram
(i.e. within a radius of $0.564^\circ$ from the indicated sight-line).
For clarity, we have imposed the additional criterion
that the photometric uncertainties
must be smaller than 0.05 mag in each band,
corresponding to a magnitude limit near $\sim$19th magnitude.
